---
title: "CAPE in using clustered transcripts in multiple tissues"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

```{r set_type}
#set testing to FALSE to run full gene expression networks and to TRUE
#to run a subset.
#testing = TRUE
testing = FALSE
is.interactive = FALSE
#is.interactive = TRUE
net.type = "signed"
#net.type = "unsigned"

args <- commandArgs(trailingOnly=T)
tissue.type <- args[1]
print(tissue.type)


if(is.na(tissue.type)){
  tissue.type = "Islet"
}
```

The purpose of this workflow is to cluster bulk tissue gene expression 
from the DO mice with WGCNA and CoExpNets. Downstream analyses will
compare the clustering from these two methods.

This workflow cluster transcripts separately in the specified tissue,
and runs Islet by default. Other possible tissues are Adipose, Liver,
and SkeletalMuscle.


```{r load_code}
library("here")
all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r load_libraries, message = FALSE, warning = FALSE, error = FALSE}
all.packages <- c("pheatmap", "gprofiler2", "CoExpNets")
load_libraries(all.packages)
```


```{r read_data}
exp.file <- here("Data", "RDS_datasets_tissues", paste0(tissue.type, ".RDS"))
tissue.exp <- readRDS(exp.file)
```

## Filter Expression
Filter the expression matrices to include only genes that have at least
a minimum amount of expression. 

Here we use the raw expression matrix to select transcripts from the rank Z
normalized expression matrix. 

```{r adjust_covar}
min.mean = 10

tissue.trans <- which(colMeans(tissue.exp$data$raw) > min.mean)
tissue.expr <- tissue.exp$data$rz[,tissue.trans]
tissue.covar <- tissue.exp$covar.matrix
adj.expr <- adjust(tissue.expr, tissue.covar)
```


Also collect clinical traits.

```{r pheno}
pheno <- read.csv(here("Data", "DO_clinical_phenotypes.csv"), stringsAsFactors = FALSE, row.names = 1)
#subset to numeric traits
num.pheno <- apply(pheno[,11:30], 2, as.numeric)
rownames(num.pheno) <- rownames(pheno)
gene.table <- tissue.exp$annot.mrna
```

## Cluster gene expression using CoExpNets

Also cluster gene expression with CoExpNets. This is WGCNA plus a k-means
clustering step that redistributes genes across modules. It generally 
identifies clusters with more even size distribution and more specific
enrichments than WGCNA.

```{r coExpNets}
#Clustering takes a long time
#set fullAnnotation to F because function defaults to human.
#we will do this ourselves by hand.
clust.dir <- here("Results", "CoExpNet_Clusters", tissue.type)
if(!file.exists(clust.dir)){dir.create(clust.dir)}
net.file <- get.files(clust.dir, want = c("net", "rds"), dont.want = "pdf", 
full.names = TRUE, ignore.case = FALSE)
if(length(net.file) == 0){
net.file = CoExpNets::getDownstreamNetwork(tissue=tissue.type,
		net.type = net.type, debug=testing, expr.data = adj.expr,
		job.path= clust.dir, save.plots = TRUE, fullAnnotation = F)
}
co.exp.net <- readRDS(net.file)
```


## Characterize Modules {.tabset .tabset-fade .tabset-pills}

In each tissue we characterized the modules using gprofiler2. We looked 
for enrichment in all GO domains, as well as KEGG and REACTOME pathways.

The following heatmaps show enriched terms for each set of modules.
Shown are the top 10 most significant terms with fewer than 500 genes.
The islet enrichment plot only shows the top 5 terms per group, since
there are so many modules in this tissue.

```{r get_module_genes}
module_genes <- function(net.obj){
  modules <- gsub("ME", "", colnames(net.obj$MEs))
  moduleV <- net.obj$moduleColors
  module.genes <- lapply(modules, function(x) names(moduleV)[which(moduleV == x)])
  names(module.genes) <- modules
  return(module.genes)
}

characterize_module <- function(enrich.table, n.terms = 10, 
pval_thresh = 0.05, max.term.size = NULL, order.by = "p_value",
decreasing = FALSE){

  low.info.words <- c("to", "of", "the", "process", "or", "by", "in", "and")

  if(length(enrich.table) == 0){return("none")}
  
  if(length(enrich.table) > 1){
    enrichments <- enrich.table[[1]]
    if(length(max.term.size) > 0){
      term.locale <- which(enrichments[,"term_size"] <= max.term.size)
    }else{
      term.locale <- 1:nrow(enrichments)
    }
    if(length(pval_thresh) > 0){
      sig.locale <- which(enrichments[,"p_value"] <= pval_thresh)
    }else{
      sig.locale <- 1:nrow(enrichments)
    }
    take.locale <- intersect(term.locale, sig.locale)
    all.terms <- enrichments[take.locale,"term_name"]
    if(!is.null(order.by)){
      ordered.terms <- enrichments[order(enrichments[,order.by], decreasing = decreasing),]
    }else{
      ordered.terms <- enrichments
    }
    all.terms <- ordered.terms[,"term_name"]
    term.words <- unique(unlist(strsplit(all.terms, " ")))
    pruned.words <- setdiff(term.words, low.info.words)
    top.words <- pruned.words[1:n.terms]
    mod.description <- paste(top.words, collapse = "-")
    return(mod.description)
  }
}
```

### Liver Module Enrichment

```{r liver_enrichment, fig.height = 25, fig.width = 7}
liver.module.genes <- module_genes(liver.net)
liver.enrich.file <- here("Results", "Expr_Clusters", "liver", "group_enrichment.RDS")
if(!file.exists(liver.enrich.file)){
  liver.enrich <- lapply(liver.module.genes, function(x) gost(x, organism = "mmusculus",
  sources = c("GO", "KEGG", "REACTOME")))
  saveRDS(liver.enrich, liver.enrich.file)
}else{
  liver.enrich <- readRDS(liver.enrich.file)
}
liver.mod.names <- sapply(liver.enrich, 
function(x) characterize_module(x, n.terms = 5, order.by = "p_value", max.term.size = 500))
#pdf("~/Desktop/liver.enrichment.pdf", width = 7, height = 25)
plot.enrichment.group(liver.enrich, max.term.size = 500, plot.label = "Liver Modules",
transformation = "sqrt", sort.by = "p_value")

#dev.off()
```

### Adipose Module Enrichment

```{r adipose_enrichment, fig.height = 20, fig.width = 7}
adipose.module.genes <- module_genes(adipose.net)
adipose.enrich.file <- here("Results", "Expr_clusters", "adipose", "adipose.enrich.RDS")
if(!file.exists(adipose.enrich.file)){
  adipose.enrich <- lapply(adipose.module.genes, function(x) gost(x, organism = "mmusculus",
  sources = c("GO", "KEGG", "REACTOME")))
  saveRDS(adipose.enrich, adipose.enrich.file)
}else{
  adipose.enrich <- readRDS(adipose.enrich.file)
}
adipose.mod.names <- sapply(adipose.enrich, 
function(x) characterize_module(x, n.terms = 5, order.by = "p_value", max.term.size = 500))
#pdf("~/Desktop/adipose.enrichment.pdf", width = 7, height = 25)
plot.enrichment.group(adipose.enrich, max.term.size = 500, 
plot.label = "Adipose Modules", transformation = "sqrt")
#dev.off()
```

### Islet Module Enrichment

```{r islet_enrichment, fig.height = 25, fig.width = 7}
islet.module.genes <- module_genes(islet.net)
islet.enrich.file <- here("Results", "Expr_clusters", "islets", "islet.enrich.RDS")
if(!file.exists(islet.enrich.file)){
  islet.enrich <- lapply(islet.module.genes, function(x) gost(x, organism = "mmusculus",
  sources = c("GO", "KEGG", "REACTOME")))
  saveRDS(islet.enrich, islet.enrich.file)
}else{
  islet.enrich <- readRDS(islet.enrich.file)
}
islet.mod.names <- sapply(islet.enrich, 
function(x) characterize_module(x, n.terms = 5, order.by = "p_value", max.term.size = 500))
#pdf("~/Desktop/islet.enrichment.pdf", width = 9, height = 32)
plot.enrichment.group(islet.enrich, max.term.size = 500, plot.label = "Islet Modules",
n.terms = 5)
#dev.off()
```

## Trait-Module Correlation {.tabset .tabset-fade .tabset-pills}

## Module overlap {.tabset .tabset-fade .tabset-pills}

We were interested in comparing gene modules from the different tissues. 
Here we compare them in two ways:
  1. Jaccard index based on included genes
  2. Correlation of module eigengenes


```{r module_overlap}
get_module_expression <- function(tissue.name, gene.id){
  if(tissue.name == "liver"){
    expr.mat <- adj.liver
  }
  if(tissue.name == "adipose"){
    expr.mat <- adj.adipose
  }
  if(tissue.name == "islet" || tissue.name == "islets"){
    expr.mat <- adj.islet
  }

  gene.locale <- match(gene.id, colnames(expr.mat))
  return(expr.mat[,gene.locale])
}

get_module_eigengene <- function(tissue.name, mod.name){
  
  mod.eig.name <- paste0("ME", mod.name)

  if(tissue.name == "liver"){
    eig.net <- liver.net
    ind.names <- rownames(adj.liver)
  }
  if(tissue.name == "adipose"){
    eig.net <- adipose.net
    ind.names <- rownames(adj.adipose)
  }
  if(tissue.name == "islet" || tissue.name == "islets"){
    eig.net <- islet.net
    ind.names <- rownames(adj.islet)
  }

  mod.locale <- which(names(eig.net$MEs) == mod.eig.name)
  mod.eig <- eig.net$MEs[[mod.locale]]
  names(mod.eig) <- ind.names
  return(mod.eig)
}

names(liver.module.genes) <- paste("liver", names(liver.module.genes), liver.mod.names, sep = "-")
names(islet.module.genes) <- paste("islet", names(islet.module.genes), islet.mod.names, sep = "-")
names(adipose.module.genes) <- paste("adipose", names(adipose.module.genes), adipose.mod.names, sep = "-")
all.module.genes <- c(liver.module.genes, islet.module.genes, adipose.module.genes)
module.pairs <- pair.matrix(1:length(all.module.genes))

module.jaccard <- module.cor <- matrix(0, nrow = length(all.module.genes), ncol = length(all.module.genes))
rownames(module.jaccard) <- colnames(module.jaccard) <- rownames(module.cor) <- colnames(module.cor) <- names(all.module.genes)
for(i in 1:nrow(module.pairs)){
  mod1 <- module.pairs[i,1]
  mod2 <- module.pairs[i,2]
  mod.comp <- jaccard.ind(all.module.genes[[mod1]], all.module.genes[[mod2]])
  module.jaccard[mod1,mod2] <- mod.comp
  module.jaccard[mod2,mod1] <- mod.comp
  mod1.name <- strsplit(names(all.module.genes)[mod1], "-")
  mod2.name <- strsplit(names(all.module.genes)[mod2], "-")
  mod1.eig <- get_module_eigengene(mod1.name[[1]][1], mod1.name[[1]][2])
  mod2.eig <- get_module_eigengene(mod2.name[[1]][1], mod2.name[[1]][2])
  
  common.ind <- intersect(names(mod1.eig), names(mod2.eig))
  mod1.ind.locale <- match(common.ind, names(mod1.eig))
  mod2.ind.locale <- match(common.ind, names(mod2.eig))
    
  mod.eig.cor <- cor(mod1.eig[mod1.ind.locale], mod2.eig[mod2.ind.locale])
  module.cor[mod1, mod2] <- mod.eig.cor
  module.cor[mod2, mod1] <- mod.eig.cor
}

```

The figures below the Jaccard indices for each set of modules compared
against the modules from the other two tissues. 

```{r plot_similarities, results = "asis", fig.width = 15, fig.height = 15}
tissue.names <- c("liver", "adipose", "islet")
for(i in 1:length(tissue.names)){
  tissue.name  <- tissue.names[i]
  cat("###", tissue.name, "\n")
  tissue.locale <- grep(tissue.name, names(all.module.genes))
  sub.jaccard <- module.jaccard[tissue.locale,-tissue.locale]
  #pdf("~/Desktop/jaccard.pdf", width = 15, height = 15)
  pheatmap(sub.jaccard)
  #dev.off()
  cat("\n\n")
}
```

The figure below shows the correlation matrix for the module eigengenes. 
Interestingly, the module eigengenes tend to cluster by tissue, and not
as much by function. For example, modules labeled with "immune" tend not
to cluster near each other.


```{r mod_eig_cor, fig.width = 20, fig.height = 20}
#pdf("~/Desktop/mod.cor.pdf", width = 20, height = 20)
pheatmap(module.cor)
#dev.off()

```

## Module mapping  {.tabset .tabset-fade .tabset-pills}

Here we map each module eigengene from CoExpNets. Compare these to the 
mapping of the WGCNA modules in Keller et al. 2018. They are pretty grassy,
except for islet module skyblue3, which has a whopping QTL. I think this 
module is probably the same as Plum1 in Keller 2018.

The islet modules are the only ones we can compare to the Keller paper 
directly. We see a lot of the same signals with similar LOD scores.
We have more modules than were reported previously. We see the same
chromosome 11 hotspot spanning multiple modules. We also see a similar
pleiotropic region on the X.

### Liver Modules
```{r map_liver_mods, fig.width = 12, fig.height = 5}
liver.scan.file <- here("Results", "Expr_Clusters", "liver", "liver.scan.RDS")
if(!file.exists(liver.scan.file)){
  liver.scans <- scan1(genoprobs, liver.net$MEs)
  saveRDS(liver.scans, liver.scan.file)
}else{
  liver.scans <- readRDS(liver.scan.file)
}
par(xpd = TRUE)
mod.dist <- dist(t(liver.scans))
mod.order <- hclust(mod.dist)$order
multilod.plot(liver.scans[,mod.order], map = map, lod.thresh = 5, border.lwd = 1, 
row.names = names(liver.module.genes)[mod.order], row.name.shift = -500, 
row.text.cex = 0.7, mar = c(2,20,2,0))
par(xpd = FALSE)
#plot(liver.scans, "MEsaddlebrown", map = map)
```

### Adipose Modules

```{r map_adipose_mods, fig.width = 12, fig.height = 5}
adipose.scan.file <- here("Results", "Expr_Clusters", "adipose", "adipose.scan.RDS")

if(!file.exists(adipose.scan.file)){
  adipose.scans <- scan1(genoprobs, adipose.net$MEs)
  saveRDS(adipose.scans, adipose.scan.file)
}else{
  adipose.scans <- readRDS(adipose.scan.file)
}
mod.dist <- dist(t(adipose.scans))
mod.order <- hclust(mod.dist)$order
par(xpd = TRUE)
multilod.plot(adipose.scans[,mod.order], map = map, lod.thresh = 4, border.lwd = 1, 
row.names = names(adipose.module.genes)[mod.order], row.name.shift = -500, 
row.text.cex = 0.7, mar = c(2,20,2,0))
par(xpd = FALSE)
#plot(adipose.scans, "MEpink", map = map)
```

### Islet Modules

```{r map_islet_modules, fig.width = 12, fig.height = 10}
islet.scan.file <- here("Results", "Expr_Clusters", "islets", "islet.scan.RDS")
if(!file.exists(islet.scan.file)){
 islet.scans <- scan1(genoprobs, islet.net$MEs)
 saveRDS(islet.scans, islet.scan.file)
}else{
  islet.scans  <- readRDS(islet.scan.file)
}
mod.dist <- dist(t(islet.scans))
mod.order <- hclust(mod.dist)$order
islet.thresh <- islet.scans[,mod.order]
islet.thresh[which(islet.thresh > 8)] <- 8
par(xpd = TRUE)
multilod.plot(islet.thresh, map = map, lod.thresh = 5, border.lwd = 1, 
row.names = names(islet.module.genes)[mod.order], row.name.shift = -800, 
row.text.cex = 0.7, mar = c(2, 20, 2, 0))
par(xpd = FALSE)

#plot(islet.scans, "MEskyblue3", map = map)
#plot(islet.scans, "MEyellowgreen", map = map)
```

## Adjust clinical phenotypes for covariates  {.tabset .tabset-fade .tabset-pills}

Adjust all traits for DO wave. Keep sex as an interactive covariate.

### Before Adjusting

```{r pheno_cor}
pheatmap(cor(num.pheno, use = "pairwise.complete.obs"), 
main = "Phenotype Correlations Before Adjusting for Covariates")
```

### After Adjusting

```{r adj_pheno_cor}
adj.pheno <- adjust(num.pheno, covar[,2:ncol(covar)])
pheatmap(cor(adj.pheno, use = "pairwise.complete.obs"), 
main = "Phenotype Correlations After Adjusting for Covariates")
```

## Create "Genotype" Object and Information Table
In CAPE we use a table of information about the markers to keep track
of where the markers are positioned on the genome. When we use modules,
we lose positional information. Here we create a dummy table labeling 
the modules from each tissue as being from a different chromosome.

```{r genotypes}
common.ind <- Reduce("intersect", list(rownames(liver.net$MEs), 
rownames(adipose.net$MEs), rownames(islet.net$MEs)))
liver.ind.locale <- match(common.ind, rownames(liver.net$MEs))
adipose.ind.locale <- match(common.ind, rownames(adipose.net$MEs))
islet.ind.locale <- match(common.ind, rownames(islet.net$MEs))

liver.MEs <- liver.net$MEs[liver.ind.locale,]
colnames(liver.MEs) <- names(liver.module.genes)

adipose.MEs <- adipose.net$MEs[adipose.ind.locale,]
colnames(adipose.MEs) <- names(adipose.module.genes)

islet.MEs <- islet.net$MEs[islet.ind.locale,]
colnames(islet.MEs) <- names(islet.module.genes)

expr.geno <- cbind(liver.MEs, adipose.MEs, islet.MEs)
```

## Build CAPE objects

Include all traits. Subsets of traits can be specified
in the cape yml file.

```{r to_geno}
#This function takes a matrix of numeric values and converts
#it into a 3D genotype array with the values scaled between
#0 and 1.

add_to_geno <- function(geno.obj, matX){

  #scale to be between 0 and 1 for cape genotypes
  pos.mat <- apply(matX, 2, function(x) x+abs(min(x)))
  #boxplot(pos.mat)
  scaled.mat <- apply(pos.mat, 2, function(x) x/max(x))
  #boxplot(scaled.mat)

  #create a 3D array of the scaled PCs
  new.geno <- array(NA, dim = c(nrow(matX), ncol(geno.obj), ncol(matX)))
  dimnames(new.geno) <- list(rownames(matX), colnames(geno.obj), colnames(scaled.mat))
  for(i in 1:ncol(scaled.mat)){
    one.mat <- matrix(scaled.mat[,i], nrow = nrow(scaled.mat), ncol = ncol(geno.obj))
    new.geno[,,i] <- one.mat
  }

  aug.geno <- abind(geno.obj, new.geno, along = 3)
  return(aug.geno)
}

```

```{r subset_ind}
#Subset one of the objects to only the individuals
#with expression before we make the cape object.
#Then we don't need to do build large objects and 
#subset them later
common.ind <- intersect(rownames(expr.geno), rownames(adj.pheno))
ind.locale <- match(common.ind, rownames(adj.pheno))
sub.obj <- dataset.clinical.phenotypes
sub.obj$pheno <- adj.pheno[ind.locale,]
```

```{r cross_obj}
cape.obj <- qtl2_to_cape(cross = sub.obj, genoprobs, map, covar[,1,drop=FALSE])
data.obj <- cape.obj$data_obj
geno.obj <- cape.obj$geno_obj

#replace data.obj$pheno with the numeric phenotype matrix and covariates
data.obj$pheno <- cbind(adj.pheno[ind.locale,], covar[ind.locale,1,drop=FALSE])

#scale expression data and add to the genotype object
new.geno.obj <- add_to_geno(geno.obj, expr.geno)

#add the information for the new "genotypes" to the data object
data.obj$geno_names <- dimnames(new.geno.obj)
names(data.obj$geno_names) <- c("mouse", "allele", "locus")

#add pseudo-chromosome data to data.obj
#cbind(liver.MEs, adipose.MEs, islet.MEs)
new.chr <- c(rep(1, ncol(liver.MEs)), rep(2, ncol(adipose.MEs)), rep(3, ncol(islet.MEs)))
new.pos <- c(1:ncol(liver.MEs), 1:ncol(adipose.MEs), 1:ncol(islet.MEs))
data.obj$chromosome <- c(data.obj$chromosome, new.chr)
data.obj$marker_location <- c(data.obj$marker_location, new.pos)
data.obj$marker_num <- 1:length(data.obj$chromosome)


#if we only want to include expression values, take out the actual genotypes
if(!include.real.genotypes){
  true.geno.locale <- 1:dim(geno.obj)[3]
  expr.geno.locale <- setdiff(1:dim(new.geno.obj)[3], true.geno.locale)
  data.obj$chromosome <- data.obj$chromosome[expr.geno.locale]
  data.obj$marker_location <- data.obj$marker_location[expr.geno.locale]
  data.obj$marker_num <- 1:length(data.obj$chromosome)
  #also reduce the "genotypes" to just two alleles.
  #we don't need all 8 if we're just looking at expression
  new.geno.obj <- new.geno.obj[,1:2,expr.geno.locale]
  data.obj$geno_names[[3]] <- data.obj$geno_names[[3]][expr.geno.locale]
  data.obj$geno_names[[2]] <- data.obj$geno_names[[2]][1:2]
}

saveRDS(data.obj, here("Data", paste0("cape_data_", geno.type, ".RDS")))
saveRDS(new.geno.obj, here("Data", paste0("cape_geno_", geno.type, ".RDS")))
```

Now run 2.cape_run_expression.Rmd.
Use the same data set name as set here.
