---
title: "Compare WGCNA and CoExpNets"
author: Anna L Tyler
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    collapsed: no
    toc: yes
    toc_float: yes
---

```{r set_type}
#set testing to FALSE to run full gene expression networks and to TRUE
#to run a subset.
#testing = TRUE
testing = FALSE
is.interactive = FALSE
#is.interactive = TRUE

tissues = c("Islet", "Liver", "Adipose", "SkeletalMuscle")

args <- commandArgs(trailingOnly=T)
tissue.idx <- as.numeric(args[1])
tissue.type <- tissues[tissue.idx]

if(is.na(tissue.type)){
  tissue.type = "Islet"
}


library(here)
wgcna.dir <- here("Results", "WGCNA_Clusters", tissue.type)
coexp.dir <- here("Results", "CoExpNet_Clusters", tissue.type)
```

## Introduction

The purpose of this workflow is to compare clusters made by WGCNA and
CoExpNets. We compare clusters fron a single tissue type based on 
gene membership, correlation of module eigengenes, functional enrichments,
and mapping.

This report is for `r tissue.type`.

```{r load_code}
all.fun <- list.files(here("Code"), full.names = TRUE, pattern = ".R")
for(i in 1:length(all.fun)){source(all.fun[i])}
```

```{r load_libraries, message = FALSE, warning = FALSE, error = FALSE}
all.packages <- c("pheatmap", "gprofiler2", "CoExpNets", "qtl2", "WGCNA")
load_libraries(all.packages)
```


```{r read_clustering_results}
wgcna.gene.mem <- readRDS(file.path(wgcna.dir, "Module_Membership.RDS"))
coexp.gene.mem <- readRDS(file.path(coexp.dir, "Module_Membership.RDS"))
```

## Basic Stats

The following plots show basic statistics of the modules. How many there
are, the number of genes, etc.

```{r compare_stats, fig.width = 4, fig.height = 4}
num.wgcna <- length(wgcna.gene.mem)
num.coexp <- length(coexp.gene.mem)
barplot(c("WGCNA" = num.wgcna, "CoExpNets" = num.coexp), main = "Number of Modules")
```


```{r compare_size, fig.width = 8, fig.height = 4}
wgcna.genes <- sapply(wgcna.gene.mem, length)
coexp.genes <- sapply(coexp.gene.mem, length)

barplot(sort(wgcna.genes), main = "WGCNA genes per Module", las = 2)
barplot(sort(coexp.genes), main = "CoExpNets genes per Module", las = 2)

#boxplot(list("WGCNA" = wgcna.genes, "CoExpNets" = coexp.genes))
```

## Jaccard Indices 

The following plots show Jaccard indices for gene membership across 
clustering methods.

```{r jaccard, fig.width = 8, fig.height = 6}
jaccard.mat <- matrix(NA, nrow = length(wgcna.gene.mem), ncol = length(coexp.gene.mem))
rownames(jaccard.mat) <- names(wgcna.gene.mem)
colnames(jaccard.mat) <- names(coexp.gene.mem)
for(i in 1:length(wgcna.gene.mem)){
  for(j in 1:length(coexp.gene.mem)){
    j.ind <- jaccard.ind(wgcna.gene.mem[[i]], coexp.gene.mem[[j]])
    jaccard.mat[i,j] <- j.ind
  }
}
pheatmap(t(jaccard.mat))
```

## Module Eigengene Correlation

The following plot shows the correlations between module eigengenes.

```{r mod_eig_cor, fig.width = 8, fig.height = 6}
wgcna.eig <- readRDS(file.path(wgcna.dir, "Module_Eigengenes.RDS"))
coexp.eig <- readRDS(file.path(coexp.dir, "Module_Eigengenes.RDS"))

cor.mat <- matrix(NA, nrow = ncol(wgcna.eig), ncol = ncol(coexp.eig))
rownames(cor.mat) <- colnames(wgcna.eig)
colnames(cor.mat) <- colnames(coexp.eig)
for(i in 1:ncol(wgcna.eig)){
  for(j in 1:ncol(coexp.eig)){
    eig.cor <- cor(wgcna.eig[,i], coexp.eig[,j])
    cor.mat[i,j] <- eig.cor
  }
}
pheatmap(t(cor.mat))
```

## Trait Correlations

The following plot shows the correlations of the module eigengenes from each
clustering method with the clinical trait.

```{r trait_cor}
exp.file <- here("Data", "RDS_datasets_tissues", paste0(tissue.type, ".RDS"))
tissue.exp <- readRDS(exp.file)
tissue.covar <- tissue.exp$covar.matrix
pheno <- read.csv(here("Data", "DO_clinical_phenotypes.csv"), stringsAsFactors = FALSE, row.names = 1)
num.pheno <- apply(pheno[,11:30], 2, as.numeric)
rownames(num.pheno) <- rownames(pheno)
adj.pheno <- adjust(num.pheno, tissue.covar)

wgcna.cor <- sapply(1:ncol(wgcna.eig), function(x) apply(adj.pheno, 2, function(y) cor(y, wgcna.eig[,x], use = "pairwise.complete.obs")))
coexp.cor <- sapply(1:ncol(coexp.eig), function(x) apply(adj.pheno, 2, function(y) cor(y, coexp.eig[,x], use = "pairwise.complete.obs")))

plot.grouped.boxes(list(t(wgcna.cor), t(coexp.cor)), 
group.labels = c("WGCNA", "CoExpNet"), type = "matrix", plot.type = "strip")
```

The following plot shows the bext positive and negative correlations
for each trait and each type of clustering. They are incredibly similar.

```{r max_cor_compare, fig.width = 10, fig.height = 6}
max.wgcna.cor <- apply(wgcna.cor, 1, function(x) c(max(x), min(x)))
max.coexp.cor <- apply(coexp.cor, 1, function(x) c(max(x), min(x)))

max.order <- order(max.wgcna.cor[1,])

plot.new()
plot.window(xlim = c(1,nrow(wgcna.cor)), ylim = c(-1, 1))
points(max.wgcna.cor[1,max.order], type = "b", col = "#beaed4", pch = 16)
points(max.wgcna.cor[2,max.order], type = "b", col = "#beaed4", pch = 16)
points(max.coexp.cor[1,max.order], type = "b", col = "#7fc97f", pch = 16)
points(max.coexp.cor[2,max.order], type = "b", col = "#7fc97f", pch = 16)
axis(2)
par(xpd = TRUE)
text(x = 1:nrow(wgcna.cor), y = rep(-1, nrow(wgcna.cor)), 
labels = colnames(num.pheno)[max.order],srt = 90, adj = 1)
par(xpd = FALSE)
abline(h = c(0, 0.5, -0.5), lty = c(1,2,2), col = "gray")
legend("topleft", legend = c("WGCNA", "CoExpNets"), pch = 16, col = c("#beaed4", "#7fc97f"))
```

## Compare Mapping

```{r qtl_fun}
summarize_qtl <- function(qtl.table){
  u_traits <- unique(qtl.table[,1])
  num.traits <- length(u_traits)
  num.qtl <- sapply(u_traits, function(x) length(which(qtl.table[,1] == x)))
  max.lod <- sapply(u_traits, function(x) max(as.numeric(qtl.table[which(qtl.table[,1] == x),"lod"])))
  qtl.chr <- table(unlist(lapply(u_traits, function(x) unique(qtl.table[which(qtl.table[,1] == x),"chr"]))))
  results <- list("Traits_with_Sig_QTL" = u_traits, "QTL_Per_Trait" = num.qtl,
  "Max_LOD" = max.lod, "QTL_Chr" = qtl.chr)
  return(results)
}
```

```{r raw_qtl, fig.width = 4, fig.height = 4}
wgcna.qtl <- readRDS(file.path(wgcna.dir, "qtl.scan.RDS"))
coexp.qtl <- readRDS(file.path(coexp.dir, "qtl.scan.RDS"))

all.wgcna.vals <- as.vector(wgcna.qtl)
all.coexp.vals <- as.vector(coexp.qtl)

boxplot(list("WGCNA" = all.wgcna.vals, "CoExpNets" = all.coexp.vals),
ylab = "LOD")

#stripchart(list("WGCNA" = all.wgcna.vals, "CoExpNets" = all.coexp.vals),
#method = "stack", pch = 16, col = "gray", vertical = TRUE)
```

```{r qtl}
wgcna.qtl <- as.matrix(read.table(file.path(wgcna.dir, "QTL.Significant.Peaks.txt"), 
header = TRUE, stringsAsFactors = FALSE))
wgcna.summ <- summarize_qtl(wgcna.qtl)

coexp.qtl <- as.matrix(read.table(file.path(coexp.dir, "QTL.Significant.Peaks.txt"), 
header = TRUE, stringsAsFactors = FALSE))
coexp.summ <- summarize_qtl(coexp.qtl)

boxplot(list("WGCNA" = as.numeric(wgcna.qtl[,"lod"]), "CoExpNets" = as.numeric(coexp.qtl[,"lod"])))
```

## Enrichment

The following plots compare enrichment statistics for modules created by the 
two methods.

```{r enrich_fun}
summarize_enrich <- function(enrich.list){
  num.with.enrichment <- length(which(sapply(enrich.list, length) > 0))
  percent.with.enrichment <- num.with.enrichment/length(enrich.list)
  terms.per.enrichment <- sapply(enrich.list, function(x) if(length(x) > 0){nrow(x[[1]])}else{0})
  result  <- list("percent.with.enrichment" = percent.with.enrichment, 
  "num.terms" = terms.per.enrichment)
  return(result)
}

```

```{r enrichment}
wgcna.enrich <- readRDS(file.path(wgcna.dir, "enrich.RDS"))
coexp.enrich <- readRDS(file.path(coexp.dir, "enrich.RDS"))

wgcna.enrich.summary <- summarize_enrich(wgcna.enrich)
coexp.enrich.summary <- summarize_enrich(coexp.enrich)

barplot(c("WGCNA" = wgcna.enrich.summary[[1]], "CoExpNets" = coexp.enrich.summary[[1]]),
main = "Percent Modules with Enrichments")

par(mfrow = c(1,2))
boxplot(list("WGCNA" = wgcna.enrich.summary[[2]], 
"CoExpNets" = coexp.enrich.summary[[2]]))

stripchart(list("WGCNA" = wgcna.enrich.summary[[2]], 
"CoExpNets" = coexp.enrich.summary[[2]]), vertical = TRUE, method = "stack",
pch = 16, col = "gray")
mtext(side = 3, text = "Number of Significantly Enriched Terms", outer = TRUE,
line = -2.5)
```